---
title: "R on Amazon EC2"
author: "Jay Emerson, Susan Wang, and Dan Snyder"
date: "4/17/2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


**Jay and Susan,
Department of Statistics, Yale University, 2013 (original version);
2016 (revised); February 2017 (revised again)**


This guide is divided into several sections.
The first section should help you create an EC2 instance on your
own from one of the Amazon basic images.  The second section
describes customizations related to R. 
This process is essentially what
[Bioconductor (link)](http://www.bioconductor.org/help/bioconductor-cloud-ami/)
would have done in creating its image(s) with customizations.  It may be
there are some images for `R` and `shiny-server` that are ready to go on
EC2, but we haven't looked recently.

Finally, we show how to take this basic process to create an instance for
the purpose of building an on-demand cluster (not checked in 2016 or
2017 revisions).  With only a few extra steps
this cluster can be used with R's `foreach` (via `doSNOW`) for parallel
computing.


## Step-by-step instructions for setting up a (possibly free) T1 Micro instance

1. Sign up for an account (if you haven't)!  Set up the account for whatever
region is appropriate (I used something in the Eastern USA).  Start here:
[http://aws.amazon.com/console/](http://aws.amazon.com/console/).

2. Create a security group. From the `EC2 Dashboard` choose `Security Groups` under
`Network and Security`.  I added a group with a VPC, and added rules for 
inbound traffic. Make the obvious choices (SSH and HTTP on ports 22 and 80).  
If you will be using Shiny, also open up port 3838 and (possibly) 10187 for 
default use of parallel programming tools (see below).

3. From the `EC2 Dashboard` under `Instances` choose `Launch Instance`.
My account is in the US East (N. Virginia) region. This will begin the 
process of creating the virtual machine you will log into.

4. Select the AMI (Amazon machine image) you would like to launch. I 
selected `Ubuntu Server 16.04 LTS` with the 64-bit version.
This is eligible for the free tier model.  It is bare-bones; choose
something bigger if you need it.

5. I'll create one (1) instance of the free tier eligible `T2 Micro` instance
type.  Don't press the blue `Review and Launch` button yet... be careful here.
Then choose `Next: Configure Instance Details`. In `Step 3: Configure Instance` 
set `Auto-assign Public IP` to Disable. This will allow you to assign an Elastic 
IP address after launch. In `Step 6: Configure Security Group` you should select 
the security group you created earlier. Now you are ready to go.

6. After selecting the correct security group for your instance, click the blue 
"Launch" button in the lower right corner of the page. If you haven't created a 
key pair yet, choose create a public/privite key pair option. These are the keys 
to the server. If you lose them, you'll be locked out. Download the key file 
and keep it in your `~/.ssh` directory (on Linux / Mac OS X systems). If you already 
created a key pair, select the "Choose an existing key pair" option. Next, click 
`Launch Instances`! After this step, there will be links that direct you to areas 
of the dashboard where you can set up billing alerts or monitor the status of your 
instance. The instance should report it as launching pretty quickly. You can watch 
from `EC2 Dashboard`.  Wait until the state is given with a green circle as `running` and
it indicates the initialization is complete.

7. If the public IP is automatically assigned, the address will change after every 
launch. We would like to assign the same public IP address to our instance every time 
it is launched. So we will create an Elastic IP address: an IP address that can
be reassigned to different EC2 instances. Go to `Network & Security` and click on 
`Elastic IPs`. Click `Allocate New Address` and follow the steps to pull an IP 
address. In the next step we will attach this address to your instance's network 
interface `eth0`. The name `eth0` is the Linux term for the primary internet 
connection to the computer.

8. Next go to `Network Interfaces` under `Networking and Security`. If your instance
launched, you should see the network interface for your instance listed here. This is
the interface to which we will bind the Elastic IP address. With the desired network
interface selected, choose `Associate Address` from `Actions` menu at the top of the 
console. The first drop down should make your Elastic IP address available for 
association with your instance's network interface. The `resassociate` check box 
allows for this assignment to override any previous assignments (because the Elastic 
IP can only associate to one network interface at a time)

9. Optionally, on your local computer, we will configure an alias for the ec2 instance. 
Doing this will make it much easier to connect to your instance. Say that your instance's 
public IP address (the Elastic IP address you assigned to your intance's network interface) 
is 19.20.21.22. Also, assume that your public key file is `ec2yale.pem`. In the `~/.ssh/config` 
file, add an entry that looks like this (using a plain text editor such as vim).

```{r eval=FALSE}
Host ec2
    Hostname 19.20.21.22
    IdentityFile ~/.ssh/ec2yale.pem
    User ubuntu
```

10. To connect to your instance, you'll use a command line interface. 
Log into the instance using the `ssh` command (secure shell). On your local machine 
(Mac OS X or Linux), ensure that your key file `ec2yale.pem` is in the `~/.ssh/` 
directory. If you lost the key file, you will need to terminate the instance and 
start over. The key file permission must give access to your user only. Key file 
permissions should be set to 400 (user execute, hidden to all others). To ensure 
this configuration, execute: `chmod 400 ec2yale.pem`. The two trailing zeros in `400` 
signify that the key is not accessible to the group or the world. If you did not set up
the host alias described in the previous step, the log in ssh command will look like:

```{r eval=FALSE}
ssh -i ~/.ssh/ec2yale.pem ubuntu@19.20.21.22
```

If you've set up the ssh config file specifying alias name `ec2`, the command will be:

```{r eval=FALSE}
ssh ec2
```

You may be asked if you are sure you want to connect to the remote machine at hte IP address specified. To allow this, type `yes`.

Windows?  You'll need a Windows SSH/SFTP client.  Check on
[Yale's software page (link)](http://software.yale.edu/).  You'll need to
set it up to use this same keyfile.  I've never tested this, so if
it doesn't work or of you discover something that should be added
to this document, please let me know.

11. Voila!  We're in.


## A little about Linux

In this context, everything you do will be at the command line, typing.
Your mouse is useless.  You don't need to know much to pull this off.
Some basic navigation:

    pwd
    ls
    ls -lat
    mkdir Stat662
    ls -lat
    cd Stat662
    pwd
    ls
    cd ..
    pwd
    ls

A [command reference (link)](http://www.ss64.com/bash/) that some have found
useful.  Editing a text file or a script?  I've pointed students at
the following [introduction to vi (link)](http://heather.cs.ucdavis.edu/~matloff/UnixAndC/Editors/ViIntro.html),
although some students also use emacs, with a [Yale help page (link)](http://zoo.cs.yale.edu/classes/cs210/help/emacs.html)
or something from [Colorado State CS (link)](http://www.cs.colostate.edu/helpdocs/emacs.html).
I should really add more to this section.


## Customization of the instance

We provide several scripts: a shell script and some R scripts.  These 
basically install R and add-on packages.  The scripts sets up Shiny
server from R Studio; this part of the installation might require
that you hit `return` at some point.

To do this customization, SSH into the instance, and pick up
the shell script (if you want to run my version without modification).
Otherwise, edit `InstallR.sh` to use your own R scripts (at your own risk).
Some possible edits would be the flags at the beginning of the file that
say whether or not to install *extras* and *mysql*. The extras include
shiny server (web hosting and data visualization) and foreach (parallel 
computing). The mysql option will allow you to setup a database server.

    instance-prompt$ sudo su
    instance-prompt$ wget http://www.stat.yale.edu/~jay/EC2/InstallR.sh
    instance-prompt$ chmod +x InstallR.sh
    instance-prompt$ ./InstallR.sh

Once this is complete, you're good to go.  On the way you may need to answer
'y' or 'Y' or something, or press `Return` on request.  Just do it.


## Testing the Web Server and Shiny

The main web server is at your IP address (here's an example):

    http://54.210.104.9

and may be customized by editing `/var/www/html/index.html`.
Shiny Server should be running, try visiting

    http://54.210.104.9:3838

You should see a working Shiny app in the top right, and a Shiny Doc
in the bottom right.  The folder for Shiny apps seems to be `/src/shiny-server`,
so if you use `sftp` to move a folder (say, `ESG` for my example)
there you can run it at

    http://54.210.104.9:3838/ESG/




## Building an on-demand cluster for parallel computing

This was not checked as of the 2016 revision.

This section simply extends material discussed above.  You can build
a compute cluster (CC) of any number of nodes of a certain type.  Amazon's
CC1 instance is 4 cores, and a CC2 instance is 8 cores (CHECK, EXPAND).
Options may be limited depending on the Amazon region (we are using
`N. Virginia`.

1. On your Dashboard, click `AMIs` under `Images`. Modify the filter for public
images, and search for `ami-0745d26e`.  Select it, and launch.  Choose your
instance type (we use `cc1.4xlarge` which may be 4 core, 23 GB RAM, with
over 800 GB of dedicated -- not shared -- disk space).  You
need to specify a placement group; we called ours `CC1-cluster`.  Click through
stuff as above (including your key pair), specify your security group, and
launch.

2. Back in the dashboard, click `Volumes` under `Elastic Block Store`.
Create a 200 GB standard volume. Be sure that the volume resides in the same
`Availability Zone` as the instance. Attach the volume to the instance, picking
`dev/sdh` as the mount point.  This will be shared across your cluster.

3. Login and sudo su.

4. Format and mount (note that the "s" in the path has been replaced by "xv")
> mkfs.ext4 /dev/xvdh  
> mkdir -m 000 /vol  
> echo "/dev/xvdh /vol auto noatime 0 0" | sudo tee -a /etc/fstab  
> sudo mount /vol  

5. Install stuff (InstallR.sh) as explained above.

6. Deal with SSH keys.  To do this, do __not__ be logged in as root! 
The command `sudo su` logs you in as root user. Files created while 
logged in as root user will not have the correct file permissions. You 
are user `ubuntu`.  Just hit `return` when asked any question about a 
passphrase.
> su ubuntu  
> cd  
> ssh-keygen -t dsa  
> cat ~/.ssh/id_dsa.pub >> ~/.ssh/authorized_keys  
> chmod 644 ~/.ssh/authorized_keys

7. Go back to EC2 Dashboard, click on instance and under Actions, select "Create Image". 

8. Maybe name your image something like `CC1-image` and then `Yes, Create`.  This will
add the image to `Images` under `AMIs`, and will take a little while (rebooting your
master instance in the process).

9. Select the image, click Launch. Select the CC1 instance type with quantity 1. 
If we wanted, say, 10 slaves, then we would pick 10 here. Make sure this instance shares the
same security group and placement group as the master.

10. Log into the master, and copy the `Public DNS` names of the slave machines into a new
file `/vol/nodelist`, one per line.  Do not include the master.

11. Edit Security Group to have port 10187 opened (could reconsider this later).

13. Fire up R. We will now do the same sort of exercise as before, making use of 
8 cores, 4 of which reside in the master "localhost", and 4 others that come from
the slave in the "nodelist" file. 

Here, we show a simple calculation done in parallel via `doSNOW` with 8 cores on the
2 nodes:

    library(doSNOW)
    library(itertools)
    setDefaultClusterOptions(port=10187)
    machines <- readLines("/vol/nodelist")
    machines <- rep(c("localhost", machines), each = 4)
    cl <- makeCluster(machines, type = "SOCK")
    registerDoSNOW(cl)
    N <- 1e6
    a <- matrix(1:N, nrow=4)
    system.time({
        b <- foreach(j=isplitIndices(ncol(a), chunks=length(machines)),
                     .combine=c) %dopar% (apply(a[,j], 2, sum))
    })
    stopCluster(cl)

## Terminating your cluster

This is important, otherwise you'll get billed about $2.50 an hour for these two
nodes!  In the Dashboard, select all the machines and terminate.  Alternatively,
you may just `stop` and we're trying to figure out what the billing implications are
of having this stuff stopped (not running) but available.  This could be a nice
option if you use this regularly.


